---
title: "How to use _skyscapeR_"
author: "Fabio Silva"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 5)
```
_skyscapeR_ is a open source R package for data reduction, visualization and analysis in skyscape archaeology, archaeoastronomy and cultural astronomy. It is intended to become a fully-fledged, transparent and peer-reviewed package offering a robust set of quantitative methods while retaining simplicity of use.


Currently it includes functions to transform horizontal (Az/Alt) to equatorial (Dec/RA) coordinates, create or download horizon profiles, plot them and overlay them with visible paths of common celestial objects/events for prehistoric or historic periods. It also includes the ability to easily construct azimuth polar plots and declination curvigrams. Future versions will add data reduction, significance testing and model selection facilities, as well as an easy-to-use Graphical User Interface.

## First Steps
### Installation
The package requires that the latest version of R is installed first. See the [R Project website](https://www.r-project.org/) for details. Also suggested is the installation of [RStudio](https://www.rstudio.com/). With R installed, install package _devtools_ by running:

```
install.packages('devtools')
```

The latest version of _skyscapeR_, and associated requirements, can then be downloaded and installed by running:
```
devtools::install_github('f-silva-archaeo/skyscapeR')
```

Upon succesful completion you should see a line saying ```* DONE (skyscapeR)```. If this doesn't happen then there should be an error message explaining what went wrong.


### Initialization
Every time you want to use the package within R it needs to be loaded. For this you need to type:
```{r}
library(skyscapeR)
```


### Data Sets and Help
The current version of _skyscapeR_ comes with a couple of datasets that can be used for learning and testing the package. These can be loaded by typing:
```
data(RugglesRSC)
```
or
```
data(RugglesCKR)
```

If you want to know more about these datasets then use the helpful ```?``` command which opens up the help page related to whatever you are asking help for.
In this case, if you want to know more about the _RugglesRSC_ dataset you can type in the console 
```
?RugglesRSC
```

This opens the manual page for the dataset, or function. In _RStudio_ this opens on the bottom-right pane by default.


### Hello, Curvigram!
Let's now do your first curvigram with _skyscapeR_. The function for this is aptly called ```curvigram()```. Go ahead and do ```?curvigram``` right now to learn more about it, including an example at the bottom of the manual page, which we will now do:
```{r}
data(RugglesRSC)
curv <- curvigram(RugglesRSC$Dec, 2)
```

This creates a curvigram based on the declination data in the _RugglesRSC_ dataset, and using an uncertainty of 2ยบ for all measurements. You can visualize it by typing:
```{r fig.show='hold'}
plotCurv(curv)
```

Be sure to check ```?curvigram``` and ```?plotCurv``` to see what other options are available.


### Adding Celestial Objects to the mix
Let's add some celestial targets to the curvigram on order to compare them with those frequency peaks. We need to create a _skyscapeR.object_ first. This is done with ```object()```:
```{r}
lunar <- object('moon', epoch=-2000, col='red', lty=2)
```

This creates an object that will include all standard lunar targets (currently only the lunar extremes), set for the year 1999 BCE (see section on epochs below), and these will be drawn in red colour with line type (lty) two, which is to say dashed (check out other types and options in ```?par```). Then redo the curvigram with this object:
```{r fig.show='hold'}
plotCurv(curv, lunar)
```

You can now see the southern major lunar extreme (sMjLX) and the southern minor lunar extreme (smnLX) declinations for the year 1999 BCE on the same plot.


## Handling Fieldwork Data
### Data Reduction for Compass Measurements
```
reduction.comp()
```

### Data Reduction for Theodolite Measurements
```
reduction.theod()
```

### Plotting Azimuths
In general, the use of azimuths for analysis and visualization in _skyscapeR_ is deprecated since azimuths are location-specific. It is preferable to convert all measurements to declinations and work with equatorial coordinates (see section below). However _it_skyscapeR_ does include a much requested function to create a polar plot of azimuth values. The function is ```plotAz()```:
```{r, fig.show='hold'}
az <- rnorm(30, 85, 20) # This creates 30 random azimuths
plotAz(az)
```

You can use the same _skyscapeR.object_ for this plot, but then you need to specify a single location (since azimuths are location-specific). At the moment the horizon altitude is assumed to be 0ยบ and flat as well.
```{r, fig.show='hold'}
obj <- object(c('sun','moon'), epoch=-4000, col=c('blue','red'), lty=c(2,3))
plotAz(az, objects=obj, loc=c(52,0))
```

### Converting Azimuths to Declination
```
az2dec()
```

## Dealing with Horizons


## Stars, Phases and Seasonality Plots
```
star()
```

```
star.phases()
```

```
plotPhases()
```

## Significance Testing for Curvigrams
```
nh.Uniform()
```

```
sigTest()
```

```
plotSigmas
```

## Where next for _skyscapeR_ ?


## References


# General vignette goodness

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
